---
title: "Random Forest Survival"
author: "John Ehrlinger  Cleveland Clinic"
date: '`r today <- Sys.Date();format(today, format="%B %d, %Y")`'
output:
  ioslides_presentation:
    widescreen: yes
    keep_md: true
  slidy_presentation: default
---

```{r setup, echo=FALSE, results=FALSE, message=FALSE, warnings=FALSE,error=FALSE}
library(knitr)
# set global chunk options for knitr. These can be changed in the header for each individual R code chunk
opts_chunk$set(fig.path='figs/rfs-', 
               fig.align='center',
               fig.height=3,
               fig.width=4,
               comment="", echo=FALSE, results=FALSE, message=FALSE, warning=FALSE,
               error=FALSE, dev='pdf')

# Setup the R environment
options(replace.assign=TRUE,object.size=Inf,expressions=100000,memory=Inf, width=100)

#################
# Load_packages #
#################
library(ggplot2) # Graphics engine for generating all types of plots
library(reshape2) # Used to modify the data for plotting
library(RColorBrewer) # Color schemes
library(scales) # For modifying ggplot

library(dplyr) # Better data manipulations

library(parallel)
options(mc.cores = detectCores()-1, rf.cores=detectCores()-1)

library(ggRandomForests)
library(randomForestSRC) 

library(ReporteRs)

library(haven)
library(plot3D)
library(hviPlotR)

#########################################################################
# Default computation settings
#########################################################################
theme_set(theme_bw())
event.marks <- c(1,4)
event.labels <- c(FALSE, TRUE)
strCol <- brewer.pal(3, "Set1")
strCol <- strCol[c(2,1,3)]
alpha <- .3

exportFigures = FALSE
exportPPT = FALSE

recalcForest= FALSE

time.offset = 2004 # Default to something... 

## Time units of pump_yr, usually it's in years.
time.unit = "year"

## The xaxis label, using pump_yr it is
xLabel="pump year"

## Set the x-axis range limits, based on the offset+ range of x (see xLabel.breaks further down)
## xLabel.steps indicates we are plotting a label every (xLabel.steps=1) year.
xLabel.steps = 1

## The histogram bin width, by default this is set to 1/2 the time units... so if pump_yr is in years,
## this would be a ", xBinWidth, " ", time.unit, " bins)"))+ bin (1/2 year)
xBinWidth = .25

## We want a small buffer around the x-values. By default, I set this to be 
xAxis.buffer = xBinWidth +.5
```
## Random Forest

Mature statistical "machine learning" method for

* Regression (continuous outcomes)
* Classification (categorical outcomes)
* Survival (time to event outcomes)
* Others (competing risk, unsupervised, etc.)

Optimized to minimize prediction error

Consistently outperforms other "off the shelf" methods

## Random Forest

Ensemble of decision trees

 * Democratic method
 * Individual weak learners
 * Aggregate to a strong learner

Non-parametric
  
  * No model assumptions
  * Nonlinear
  * Interactions
    
## Data

Data set has:

* $n$ observations
* $p$ independent variables

Ideally, want $n \rightarrow$ everyone (unrealistic)

Instead simulate with the Bootstrap

* Randomly select $n$ observations with replacement (b)
* On average 36.8% left out of bootstrap (oob)

## Random Forest

Grow a collection of independent decision trees

* One for each Bootstrap data set
* Test with the associated oob data set

But decision trees are 

* Inherently unstable 
* Tend to over fit training data

They are an ideal weak learner suitable for RF application

## Growing a Decision Tree

Recursively partition the data

* Split data nodes (set) into two daughter nodes
* Repeat to exhaustion

Two requirements

* Split rule
* Stopping rule 

## Growing a Decision Tree 

Split rule

Test each variable for optimal node segmenting

* Optimize over classes of categorical variables
* Optimize along values of continuous variables

Choose optimal variable

Dependent on the problem domain

* Regression - MSE
* Classification - Gini index (Generalize Binomial Variance)
* Survival - Log-rank

Optimally segregate two groups of observations

```{r tree-diagram}
# Build an artificial tree
trData <- tbl_df(data.frame(matrix(NA, ncol=5, nrow=19)))
colnames(trData)<- c("variable", "nodeID", "parentID", "depth", "terminal")

trData$terminal=FALSE

trData$nodeID <- 1:dim(trData)[1]
trData$parentID <- c(0,1,1,2,2,3,3, 4, 4, 5,  5, 7,  7,  8,  8,  9,  9, 11, 11)
trData$depth[1] <- 0
trData$x <- c(-5,-10,2,-13,-5,0,7,-17,-10,-7,-2, 5,10,-19,-15,-12,-8,-4,0)

trData$terminal[c(6,10,12,13,10,18,19,14,15,16,17)] <- TRUE

trData$variable <- trData$nodeID
trData$variable[which(trData$variable==9)] <- 3
trData$variable[which(trData$variable==11)] <- 3
trData$variable[which(trData$variable==7)] <- 8
trData$variable <- paste("V", trData$variable, sep="")
trData$variable[which(trData$terminal)] <- NA
trData$variable<- factor(trData$variable)

edgeList <- tbl_df(data.frame(matrix(ncol=4, nrow=18)))
colnames(edgeList) <- c("fromX", "toX", "fromY", "toY")

for(ind in 2:dim(trData)[1]){
  trData$depth[ind] <- trData$depth[trData$parentID[ind]] +1
}

for(ind in 2:dim(trData)[1]){
  edgeList[ind-1,] <- c(trData$x[trData$parentID[ind]], 
                        trData$x[ind],trData$depth[trData$parentID[ind]],
                        trData$depth[ind])
}

dimX <- c(-21,12)
dimY <- c(-1,5)
```

## Growing a Decision Tree


``` {r treeDiagram0}
dpth <- 0
trD <- filter(trData,depth <= dpth)
edL <- filter(edgeList,toY <= dpth)
trD$variable[which(trData$depth == dpth)] <- NA
ooD <- trD
ooD$variable <- NA
ooD$nodeID <- 0
ooD$x=5
trD <- rbind(trD, ooD)
dtree <- ggplot(trD)+
  #   geom_segment(aes(x=fromX, y=fromY, xend=toX, yend=toY), data=edL) + 
  geom_point(aes(x=x,y=depth, shape=terminal, fill=variable),
             size=15)+
  geom_text(aes(x=x,y=depth, label=c("b", "oob" )))+
  scale_shape_manual(values=c(21,22))+
  coord_cartesian(xlim=dimX, ylim=dimY)+
  scale_fill_brewer(palette = "Dark2", na.value="white")+
  coord_cartesian(xlim=dimX, ylim=dimY)+
  theme(legend.position="none",
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.ticks=element_blank(),
        axis.text=element_blank()
  )+ 
  labs(x="", y="")+
  coord_cartesian(y=dimY, x=dimX)
dtree 

```


## Growing a Decision Tree

``` {r treeDiagram1}
dpth <- 1
trD <- filter(trData,depth <= dpth)
edL <- filter(edgeList,toY <= dpth)
trD$variable[which(trData$depth == dpth)] <- NA
trD <- rbind(trD, ooD)
trD.t <- trD
trD.t$variable <- as.character(trD.t$variable)
trD.t$variable[nrow(trD.t)] <- "oob"

dtree <- ggplot(trD)+
  geom_segment(aes(x=fromX, y=fromY, xend=toX, yend=toY), 
               data=edL) + 
  geom_point(aes(x=x,y=depth, shape=terminal, 
                 fill=variable),size=15)+
  geom_text(aes(x=x,y=depth, label=variable), data=trD.t)+
  scale_shape_manual(values=c(21,22))+
  coord_cartesian(xlim=dimX, ylim=dimY)+
  scale_fill_brewer(palette = "Dark2", na.value="white")+
  theme(legend.position="none",
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.ticks=element_blank(),
        axis.text=element_blank()
  )+ 
  labs(x="", y="")
dtree +
  coord_cartesian(y=dimY, x=dimX)
```


## Growing a Decision Tree

``` {r treeDiagram4}
dpth <- 4
trD <- filter(trData,depth <= dpth)
edL <- filter(edgeList,toY <= dpth)
trD$variable[which(trData$depth == dpth)] <- NA
trD <- rbind(trD, ooD)
trD.t <- trD
trD.t$variable <- as.character(trD.t$variable)
trD.t$variable[nrow(trD.t)] <- "oob"
decTree <- ggplot(trD)+
  geom_segment(aes(x=fromX, y=fromY, xend=toX, yend=toY), 
               data=edL) + 
  geom_point(aes(x=x,y=depth, shape=terminal, 
                 fill=variable),size=15)+
  geom_text(aes(x=x,y=depth, label=variable), data=trD.t)+
  scale_shape_manual(values=c(21,22))+
  coord_cartesian(xlim=dimX, ylim=dimY)+
  scale_fill_brewer(palette = "Dark2", na.value="white")+
  theme(legend.position="none",
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.ticks=element_blank(),
        axis.text=element_blank()
  )+ 
  labs(x="", y="")
decTree +
  coord_cartesian(y=dimY, x=dimX)

```


##  Growing a Decision Tree

Stopping Rule defines Terminal Nodes

* Minimal number of members
* Homogeneity

Defaults depend on the problem domain

* Regression - min 5 unique cases
* Classification - homogeneous node (min of 1)
* Survival - min 3 unique cases

## Testing a Decision Tree

Tree sorts each observation into a unique terminal nodes

Test the tree with oob data.

  * Sort test observations into terminal nodes
  * Predict from training observations
  * Compare with test response
  
## Testing a Decision Tree 

```{r decisionTree}
decTree
```

## Decision Tree Prediction

Defined by terminal node membership. 

* Fit a model to training set members
* Predict from model

One model for each terminal node within the tree.

Depends on the problem domain

* Regression - mean value
* Classification - probability of class membership
* Survival - Kaplan--Meier estimates

## Random Forest Trees

A forest of independent decision trees

* Independent bootstrap training data
* Add extra randomization step

At each node split, RF randomly selects a subset (mtry $\le p$) of candidate variables for the split rule optimization

Default depends on the problem domain

* Regression - mtry = ceiling$(p/3)$ 
* Classification - mtry = ceiling$(\sqrt{p})$ 
* Survival - mtry = ceiling$(\sqrt{p})$ 

## Random Forest Prediction

A forest of independent decision trees

* Observations in a terminal node have the same predicted outcome
* Bagging (Bootstrap Aggregation) over all trees

Default depends on the problem domain

* Regression - average estimates
* Classification - voting or average probabilty
* Survival - average survival estimates

## Random Forest Performance

Measure of generalization error

* oob data used to calculate forest prediction error

Depends on the problem domain

* Regression - MSE
* Classification - Misclassification error
* Survival - Harrell's concordance index 

## Breiman's Two Cultures

Machine Learning vs. Statistics

Machine Learning:

* Prediction, Prediction, Prediction
* Black box modeling

Statistics:

* Why? 
* Information on underlying process

Random Forest:

* Why not both?
* Insight into the black box of prediction

## Random Survival Forest

Extension to time to event data

* Developed at Cleveland Clinic
* Grants and contracts from NHLBI


## PBC Example
Primary Biliary Cirrhosis (PBC) of the liver data set
(Fleming and Harrington 1991)

Randomized  trial of D-penicillamine (DPCA) 

Mayo Clinic

312 patients from 1974 to 1984

* 125 deaths
* 17 variables
``` {r setup-pbc}
################ Default Settings ##################
theme_set(theme_bw())     # A ggplot2 theme with white background

## Set open circle for censored, and x for events 
event.marks <- c(1, 4)
event.labels <- c(FALSE, TRUE)

## We want red for death events, so reorder this set.
strCol <- brewer.pal(3, "Set1")[c(2,1,3)] 
data(pbc, package = "randomForestSRC")

library("reshape2")        # Transforming wide data into long data (melt)

## Not displayed ##

## Set modes correctly. For binary variables: transform to logical
## Check for range of 0, 1
## There is probably a better way to do this.
for(ind in 1:dim(pbc)[2]){
  if(!is.factor(pbc[, ind])){
    if(length(unique(pbc[which(!is.na(pbc[, ind])), ind]))<= 2) {
      if(sum(range(pbc[, ind], na.rm = TRUE) ==  c(0, 1)) ==  2){
        pbc[, ind] <- as.logical(pbc[, ind])
      }
    }
  }else{
    if(length(unique(pbc[which(!is.na(pbc[, ind])), ind]))<= 2) {
      if(sum(sort(unique(pbc[, ind])) ==  c(0, 1)) ==  2){
        pbc[, ind] <- as.logical(pbc[, ind])
      }
      if(sum(sort(unique(pbc[, ind])) ==  c(FALSE, TRUE)) ==  2){
        pbc[, ind] <- as.logical(pbc[, ind])
      }
    }
  }
  if(!is.logical(pbc[, ind]) & 
       length(unique(pbc[which(!is.na(pbc[, ind])), ind]))<= 5) {
    pbc[, ind] <- factor(pbc[, ind])
  }
}
# Convert age to years
pbc$age <- pbc$age/364.24
pbc$years <- pbc$days/364.24
pbc <- pbc %>% select(-days)
pbc$treatment <- as.numeric(pbc$treatment)
pbc$treatment[which(pbc$treatment == 1)] <- "DPCA"
pbc$treatment[which(pbc$treatment == 2)] <- "placebo"
pbc$treatment <- factor(pbc$treatment)

cls <- sapply(pbc, class) 

labels <- c("Event (F = censor, T = death)", 
            "Treament (DPCA, Placebo)", 
            "Age (years)", 
            "Female = T", 
            "Presence of Asictes", 
            "Presence of Hepatomegaly", 
            "Presence of Spiders", 
            "Edema (0, 0.5, 1)", 
            "Serum Bilirubin (mg/dl)", 
            "Serum Cholesterol (mg/dl)", 
            "Albumin (gm/dl)", 
            "Urine Copper (ug/day)", 
            "Alkaline Phosphatase (U/liter)", 
            "SGOT (U/ml)", 
            "Triglicerides (mg/dl)", 
            "Platelets per cubic ml/1000", 
            "Prothrombin time (sec)", 
            "Histologic Stage", 
            "Time (years)")

dta.labs <- data.frame(cbind(names = colnames(pbc), label = labels, type = cls))
# Put the "years" variable on top.
dta.labs <- rbind(dta.labs[nrow(dta.labs),], dta.labs[-nrow(dta.labs),])

st.labs <- as.character(dta.labs$label)
names(st.labs) <- rownames(dta.labs)
```

## PBC Example

``` {r fh-model}
## Not displayed ##
# Create a table summarizing the ph model from fleming and harrington 1991
fleming.table <- data.frame(matrix(ncol = 4, nrow = 5))
fleming.table[,1] <- 
  c("Age", "log(Albumin)", "log(Bilirubin)", "Edema", "log(Prothrombin Time)")
colnames(fleming.table) <- c("Variables","Coef.", "Std. Err.", "Z stat.")
fleming.table[,2] <- c(0.0333, -3.0553,0.8792, 0.7847, 3.0157) 
fleming.table[,3] <- c(0.00866, 0.72408,0.09873,0.29913,1.02380) 
fleming.table[,4] <- c(3.84,-4.22,8.9,2.62,2.95) 

kable(fleming.table, 
      row.names=NA,
      format="markdown",
      digits = 3,
      booktabs=TRUE)
```

## Random Survival Forest

``` {r pbc-forest}
# in reality, we use data caching to make vignette 
# compilation quicker. The rfsrc_pbc forest is stored
# as a ggRandomForests data sets
#
# This code block produces the R output from the 
# rfsrc grow block above. We set the chunk argument 
# "echo=FALSE" above so this code does not show up 
# in the manuscript.
data(rfsrc_pbc, package = "ggRandomForests")
ggRFsrc <- plot.gg_rfsrc(rfsrc_pbc, alpha = .2) + 
  scale_color_manual(values = strCol) + 
  theme(legend.position = "none") + 
  labs(y = "Survival Probability", x = "time (years)")+
  coord_cartesian(y = c(-.01,1.01))

# Display the figure
show(ggRFsrc)

```

## Random Survival Forest

``` {r pbd-treatment} 
gg_plt <-  plot(gg_rfsrc(rfsrc_pbc, by="treatment")) + 
  theme(legend.position = c(.2,.2)) + 
  labs(y = "Survival Probability", x = "time (years)",
       color="Treatment", fill="Treatment")+
  scale_color_brewer(palette="Set1")+
  coord_cartesian(y = c(-.01,1.01))
gg_plt
```  

## Variable Selection

Two independent methods

Variable IMPortance (VIMP)

* Based on RF Prediction Error
* Measures the impact of variable misspecification

Minimal Depth

* Property of decision tree construction 
* Measures how a variable segments nodes

## Variable Selection - VIMP

Prediction error (PE) estimate from oob data

For each variable:

* Randomize values within the variable
* Predict with randomized data
* Calculate a New Prediction Error estimate (NPE)

VIMP = PE - NPE 

* Positive value: important in reducing error
* Near zero: no impact on prediction
* Negative value: noise variable 

## Variable Selection - VIMP
``` {r rf-pbc-vimp}
ggvm <- plot.gg_vimp(rfsrc_pbc, lbls = st.labs) + 
  theme(legend.position = c(.8,.2))+
  labs(fill = "VIMP > 0")
ggvm
```

##  Variable Selection - Minimal Depth

Within each tree

* Number the node split levels
* Find the minimum split level for each variable


##  Variable Selection - Minimal Depth
``` {r treeDepth}
decTree <- decTree+ 
  scale_y_continuous(breaks=0:4)+
  theme_bw()+
  theme(legend.position="none",
        panel.border = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.x=element_blank(),
        axis.line.y = element_line(color="black"),
        panel.grid.major.y = element_line(color="black")
  )+ 
  labs(x="", y="Tree Depth")
decTree
```

##  Variable Selection - Minimal Depth

Average minimal split levels 

  * each variable 
  * over the forest

Lower values split largest nodes

##  Variable Selection - Minimal Depth
``` {r mindepth-pbc}
data(varsel_pbc, package = "ggRandomForests")
gg_md <- gg_minimal_depth(varsel_pbc)
ggm <- plot(gg_md, lbls = st.labs)
ggm
```

## Random Forest

VIMP and Minimal Depth 

* which variables contribute to forest prediction?

Variable dependence 

* How does response depend on variables?

## Variable Dependence

Two Options:

Variable Dependence

  * Observation Based
  
Partial Dependence

  * Population Based

## Variable Dependence

Observation based

  * Predicted value for each observation
    * At selected times for survival
    
  * Against variable value

## Variable Dependence

``` {r rfsrc-plot3Mnth-pbc}
gg_plt <- ggRFsrc + 
  geom_vline(aes(xintercept = c(1, 3)), linetype = "dashed") + 
  coord_cartesian(x = c(0, 5))

gg_plt
```

## Variable Dependence

``` {r variable-plot-pbc}
gg_v <- gg_variable(rfsrc_pbc, time = c(1, 3), 
                    time.labels = c("1 Year", "3 Years"))
xvar <- c("bili", "albumin", "copper", "prothrombin", "age")

# The categorical variable
xvar.cat <- c("edema")

# panel the remaining continuous variable dependence plots.
gg_plt <- plot(gg_v, xvar = xvar, panel = TRUE, 
               se = FALSE, alpha = .4, span=1)+ 
  labs(y = "Survival") + 
  theme(legend.position = "none") + 
  scale_color_manual(values = strCol, labels = event.labels) + 
  scale_shape_manual(values = event.marks, labels = event.labels)+
  coord_cartesian(y = c(-.05,1.05))

gg_plt

```



## Partial Dependence

Population Based

  * Create nomograms for each observation
    * Across values of variable of interest
    * At selected times for survival
    
  * Average response

## Partial Dependence

``` {r rfs-points}
# Create nomograms for bili
xv <- rfsrc_pbc$xvar

bilipts <- quantile_pts(xv$bili, groups=6)
bilipts <- bilipts[-c(1, length(bilipts))]
gg_v <- gg_variable(rfsrc_pbc, time =3)

gg_plt <- plot(gg_v, xvar = "bili", 
               se = FALSE, alpha = .4, span=1)+ 
  labs(y = "Survival (3 years)", x="Bilirubin") + 
  theme(legend.position = "none") + 
  scale_color_manual(values = strCol, labels = event.labels) + 
  scale_shape_manual(values = event.marks, labels = event.labels)+
  coord_cartesian(y = c(-.05,1.05))+
  scale_x_continuous(breaks=seq(0,30,5))+
  geom_vline(aes(xintercept = bilipts), linetype = "dashed")

gg_plt
if(exportPPT){
  ggsave(gg_plt+theme_dark_ppt()+theme(legend.position="none"
  ),          
  path="../graphs",          
  file="uab-fig10.pdf", 
  width=4*2, height=3*2)
  embedFonts(file="../graphs/uab-fig10.pdf", outfile= "../graphs/uab-fig10-ft.pdf")
  
}

```

## Partial Dependence

``` {r rfs-nomo}
# Create nomograms for bili
prd <- lapply(bilipts, function(bl){
  xv$bili <- bl
  predict(rfsrc_pbc, 
          newdata = xv,
          na.action="na.impute",
          importance="none")
}) 

gg_rf <- lapply(prd, function(st){
  ss <- data.frame(st$survival)
  colnames(ss) <- as.character(rfsrc_pbc$time.interest)
  ss$ptid <- 1:nrow(ss)
  ss$cens <- rfsrc_pbc$yvar$status
  melt(ss, id.vars=c("ptid", "cens"))})

gg_rf <- lapply(1:length(bilipts), function(ind){
  gg_rf[[ind]]$bili <- paste("bilirubin =", bilipts[ind])
  gg_rf[[ind]]})

ggdt <- do.call(rbind, gg_rf)
ggdt$cens <- as.logical(ggdt$cens)
ggdt$bili <- factor(ggdt$bili, levels=unique(ggdt$bili))
ggdt$variable <- as.numeric(as.character(ggdt$variable))
ggdt$ptid <- factor(ggdt$ptid)

gg_plt <- ggplot(ggdt, aes(x=variable, y=value, color=cens, by=ptid))+
  geom_step(alpha=.2)+ 
  scale_color_manual(values = strCol, labels = event.labels) + 
  scale_shape_manual(values = event.marks, labels = event.labels)+
  coord_cartesian(y = c(-.05,1.05), x=c(-.1,5.1))+
  labs(y = "Survival", x="Time (Years)") + 
  theme(legend.position="none")+
  geom_vline(aes(xintercept =3), linetype = "dashed") + 
  facet_wrap(~bili)

gg_plt
if(exportPPT){
  ggsave(gg_plt+theme_dark_ppt(),          
  path="../graphs",          
  file="uab-fig11.pdf", 
  width=4*2, height=3*2)
  embedFonts(file="../graphs/uab-fig11.pdf", outfile= "../graphs/uab-fig11-ft.pdf")
}

```

## Partial Dependence

``` {r nomogram-vdep}
ggvd <- lapply(prd, function(st){
  indx <- which(rfsrc_pbc$time.interest >3)[1] - 1
  ss <- data.frame(yhat=st$survival[,indx])
  ss$ptid <- 1:nrow(ss)
  ss$cens <- rfsrc_pbc$yvar$status
  ss})

ggvd <- lapply(1:length(bilipts), function(ind){
  ggvd[[ind]]$bili <- bilipts[ind]
  ggvd[[ind]]})

ggmn <- data.frame(t(sapply(ggvd, function(st){
  c(st$bili[1], mean(st$yhat))
})))

ggdt <- do.call(rbind, ggvd)
ggdt$cens <- as.logical(ggdt$cens)

gg_plt <- ggplot(ggdt)+
  geom_point(aes(x=bili, y=yhat, shape=cens, color=cens),alpha=.4)+
  labs(y = "Survival (3 years)", x="Bilirubin") + 
  theme(legend.position = "none") + 
  scale_color_manual(values = strCol, labels = event.labels) + 
  scale_shape_manual(values = event.marks, labels = event.labels)+
  scale_x_continuous(breaks=seq(0,30,5))+
  coord_cartesian(y = c(-.05,1.05), x=c(-1,29))
gg_plt
```

## Partial Dependence

``` {r nomogram-vdep-mean}

gg_plt <- ggplot(ggdt)+
  geom_boxplot(aes(x=bili, y=yhat, by=factor(bili)), color="black", outlier.shape = NA)+
  geom_jitter(aes(x=bili, y=yhat, shape=cens, color=cens),alpha=.4)+
  labs(y = "Survival (3 years)", x="Bilirubin") + 
  theme(legend.position = "none") + 
  scale_color_manual(values = strCol, labels = event.labels) + 
  scale_shape_manual(values = event.marks, labels = event.labels)+
  scale_x_continuous(breaks=seq(0,30,5))+
  coord_cartesian(y = c(-.05,1.05), x=c(-1,29))
gg_plt
if(exportPPT){
  ggsave(gg_plt+theme_dark_ppt()+theme(legend.position="none"
  ),          
  path="../graphs",          
  file="uab-fig12.pdf", 
  width=4*2, height=3*2)
  embedFonts(file="../graphs/uab-fig12.pdf", outfile= "../graphs/uab-fig12-ft.pdf")
  
}
```

## Partial Dependence

``` {r pbc-partial} 
data("partial_pbc", package = "ggRandomForests")
xvar <- c(xvar, xvar.cat)
gg_dta <- mclapply(partial_pbc, gg_partial)

# Combine the timed gg_partial objects together.
pbc_ggpart <- combine.gg_partial(gg_dta[[1]], gg_dta[[2]], 
                                 lbls = c("1 Year", "3 Years"))

ggpart <- pbc_ggpart
ggpart$edema <- NULL

gg_plt <- plot(ggpart[[1]], se = FALSE) + 
  labs(x = "Bilirubin", y = "Survival", color = "Time", shape = "Time") +
  theme(legend.position = c(.8, .2)) + 
  scale_x_continuous(breaks=seq(0,30,5))+
  coord_cartesian(y = c(25,101))
gg_plt
if(exportPPT){
  ggsave(gg_plt+theme_dark_ppt()+theme(legend.position="none"
  ),          
  path="../graphs",          
  file="uab-fig13.pdf", 
  width=4*2, height=3*2)
  embedFonts(file="../graphs/uab-fig13.pdf", outfile= "../graphs/uab-fig13-ft.pdf")
  
}
```

## Partial Dependence
``` {r partialpanel}
ggpart[[1]] <- NULL
gg_plt <- plot(ggpart, se = FALSE, panel = TRUE) + 
  labs(x = "", y = "Survival", color = "Time", shape = "Time") +
  theme(legend.position = "none") + 
  coord_cartesian(y = c(25,101))
gg_plt
if(exportPPT){
  ggsave(gg_plt+theme_dark_ppt()+theme(legend.position="none"
  ),          
  path="../graphs",          
  file="uab-fig14.pdf", 
  width=4*2, height=3*2)
  embedFonts(file="../graphs/uab-fig14.pdf", outfile= "../graphs/uab-fig14-ft.pdf")
  
}
```

## Partial Dependence

```{r pbc-timeSurface}
# Restrict the time of interest to less than 5 years.
time_pts <- rfsrc_pbc$time.interest[which(rfsrc_pbc$time.interest<=5)]

# Find the 50 points in time, evenly space along the distribution of 
# event times for a series of partial dependence curves
time_cts <-quantile_pts(time_pts, groups = 50)

# Load the stored partial coplot data.
data(partial_pbc_time)

# We need to attach the time points of interest to our data.
time.tmp <- do.call(c,lapply(time_cts, 
                             function(grp){rep(grp, 50)}))

# Convert the list of plot.variable output to gg_partial
partial_time <- do.call(rbind,lapply(partial_pbc_time, gg_partial))

# attach the time data to the gg_partial_coplot
partial_time$time <- time.tmp

# Modify the figure margins to make it larger
par(mai = c(0.5,0.55,0,0))

# Transform the gg_partial_coplot object into a list of three named matrices
# for surface plotting with plot3D::surf3D
srf <- surface_matrix(partial_time, c("time", "bili", "yhat"))

# Generate the figure.
surf3D(x = srf$x, y = srf$y, z = srf$z, col = heat.colors(25),
       colkey = FALSE, border = "black", bty = "b2", 
       shade = 0.5, expand = 0.5, theta=110, phi=15,
       lighting = TRUE, lphi = -50, ticktype="detailed",
       ylab = "Bilirubin", xlab = "Time", zlab = "Survival"
)

# Extract the 1 and 3 year points.
# Find the indices of the points closest in time
t.pts <- sapply(c(1,3), function(pt){min(abs(srf$x - pt), na.rm=TRUE)})
indx <- vector("list", length=2)
indx[[1]] <- which(abs(srf$x - 1) < t.pts[1]+1.e-5)
indx[[2]] <- which(abs(srf$x - 3) < t.pts[2]+1.e-5)

# Generate curves along 1 and 3 year partial dependence 
alt <- lapply(indx, function(ind){
  lines3D(x=srf$x[ind], y=srf$y[ind],z=srf$z[ind],
          add=TRUE, col="blue", lwd=6)
})


if(exportPPT){
  ##--------
  # For each graph, addSlide. The graphs require the
  # Title and Content template.
  pdf( file="../graphs/uab-fig15.pdf", 
       width=4*2, height=3*2)
  # Modify the figure margins to make it larger
  par(mai = c(0.3,0.65,0,0.2))
  
  # Transform the gg_partial_coplot object into a list of three named matrices
  # for surface plotting with plot3D::surf3D
  srf <- surface_matrix(partial_time, c("time", "bili", "yhat"))
  
  # Generate the figure.
  surf3D(x = srf$x, y = srf$y, z = srf$z, col = heat.colors(25),
         colkey = FALSE, border = "black", bty = "b2", 
         shade = 0.5, expand = 0.5, theta=110, phi=15,
         lighting = TRUE, lphi = -50, ticktype="detailed",
         ylab = "Bilirubin", xlab = "Time", zlab = "Survival"
  )
  
  # Generate curves along 1 and 3 year partial dependence 
  alt <- lapply(indx, function(ind){
    lines3D(x=srf$x[ind], y=srf$y[ind],z=srf$z[ind],
            add=TRUE, col="blue", lwd=6)
  })
  
  dev.off()
  
  embedFonts(file="../graphs/uab-fig15.pdf", outfile= "../graphs/uab-fig15-ft.pdf")
  
}
```
